Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data

### 2021-11-02 analysis/paper_draft2.Rmd
edits - ongoing

---
title: "Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data"
date: "`r Sys.Date()`"
author:
  - Sachin S. Pandya
#    orcid_id: 0000-0001-7387-1307
  - Xiaomeng Li
  - Timothy E. Moore
#    orcid_id: 0000-0002-9576-0517
#output: distill::distill_article
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: true
    fig_width: 5
    fig_height: 4
header-includes: 
  - \usepackage{amsmath}
fontsize: 12pt
bibliography: references.bib
abstract: |
  US law bars using peremptory strikes during jury selection because of prospective juror race, ethnicity, gender, or membership in certain other cognizable classes. Here, we propose a Bayesian approach for detecting such illegal bias. We show how to use historical data on an attorney's use of peremptory strikes in past cases to estimate, in a current trial, how likely an attorney has struck prospective jurors because of cognizable-class membership. We also use the _power prior_ to adjust the weight of such historical data in the analysis. We show the utility of our model on simulated data. Finally, we extend this with a prototype software application to detect bias in peremptory-challenge in real time during jury-selection. We demonstrate the app using actual historical strike data from a convenience sample of cases from the federal district court of Connecticut.

@@

In any case, evidence of such illegal bias may include data on a party's use of peremptory challenges in past cases. (_Flowers v. Mississippi_ -@flowers2019, p. 2243; Wash. Rule 37(g)(v) -@WashR37; Calif. Code of Civil Procedure $\S$ 231.7(d)(3)(G)). Prior studies have collected historical strike data in past cases and reported the observed difference in strike rates by race or gender. Typically, they test for the probability of observing a non-zero difference in strike rates by the race or gender of the struck prospective jurors, given repeated sampling from a hypothetical population of peremptory strikes with zero such difference [e.g., @Eisenberg2017; @Grosso2012; for discussion, see @Gastwirth2014; @Gastwirth2013]. Prior studies have also modeled how much a prospective juror's race affected the odds of being struck. For different modeling approaches using the same historical strike data from Mississippi, see @Craft2018a; @DeCamp2021; and @Dunn2021.

In this paper, we extend a Bayesian approach to _Batson_ and similar challenges [@Kadane2018] to incorporate historical strike data. In this approach, we specify a model of the peremptory-strike process in the court of interest that includes a bias parameter to which we assign an initial prior distribution. Then, we use an attorney's strike data from past trials, as generated by that same peremptory-strike process, to estimate a posterior distribution for that bias parameter. In so doing, we use the _power prior_ [@Ibrahim2015] to control how much the strike history from past cases affects the posterior distribution of the bias parameter.  Finally, we use that posterior distribution as the prior for the bias parameter in the present case and update the posterior based on data on the target attorney's strikes in that present case.

@@

\begin{equation}
L(b|\delta) = L(b|\delta)(L(b|\delta_0))^{a_0}\exp(-\frac{b^2}{8})
\label{eq:model4}
\end{equation}

@@

To do this, we generated strike data for a current trial using seven different values of the bias parameter ("current" b $= \{-3,-2,-1,0,1,2,3\}$. As introduced above, $b>0$ ($b<0$) represents bias against (for) a prospective juror within (outside) the cognizable class, while $b=0$ denotes no bias.

We also generated historical strike data of three sizes (same, double, and triple the size of current data, i.e., data on one previous trial, two previous trials) using seven different values of the bias parameter for generating that historical data ("historical b" $= \{-3,-2,-1,0,1,2,3\}$).

We let historical $b$ differ from current $b$ (our parameter of interest) to simulate noise associated with historical strike data that is incomplete not at random. Alternatively, this difference simulates a real difference in the bias parameter that depends on a feature that the current trial has but that all the past trials in the historical data do not. To illustrate, suppose prosecutor with a bias value conditional on defendant race: $b > 0$ if the criminal defendant is Black, $b = 0$ if not. If our historical strike data contains only in cases with non-Black defendants, accounting for that historical data may lead us to underestimate prosecutor's bias ($b$) in the current case with a Black defendant.

@@

We considered 735 scenarios in total, i.e., the combination of 7 different bias parameters of current data and of historical data, 3 different amounts of historical data and 5 different values of the power prior weight parameter $a$. For each scenario, we generated 1000 datasets and fit the model on those datasets. We recorded the mean value of the posterior mean and credible intervals of one thousand model fits for each scenario. In turn, for three credible interval levels (80\%, 90\% and 95\%), we recorded the coverage rates, i.e., the proportion of times those intervals actually contained the true value of the (current) bias parameter $b$.

```{r b80}

@@

p80 <- ggplot()+
  geom_tile(data = sims, aes(x = d, y = d_h,
                        fill = as.factor(sig.n)),
            color="black")+
  facet_grid(h~a, labeller = label_both)+
  scale_x_continuous(breaks = c(-3:3), expand = c(0,0))+
  scale_y_continuous(breaks = c(-3:3), expand = c(0,0))+
  xlab("Current b")+
  ylab("Historical b")+
  scale_fill_manual(values = c("white", "blue"))+
  coord_equal() +
  theme_minimal() +
  labs(fill = "coverage") +
  theme(axis.text = element_text(size =10), 
        #axis.title = element_text(size = 16), 
        #legend.title.align = 0.05, 
        #legend.title = element_text(size = 16), 
        legend.text = element_text(size = 14))
#p80

@@

```{r b90}

@@

p90 <- ggplot()+
  geom_tile(data = sims, aes(x = d, y = d_h,
                        fill = as.factor(sig.n)),
            color="black")+
  facet_grid(h~a, labeller = label_both)+
  scale_x_continuous(breaks = c(-3:3), expand = c(0,0))+
  scale_y_continuous(breaks = c(-3:3), expand = c(0,0))+
  xlab("Current b")+
  ylab("Historical b")+
  scale_fill_manual(values = c("white", "blue"))+
  coord_equal()+
  theme_minimal()+
  labs(fill = "coverage") +
  theme(axis.text = element_text(size =10), 
        axis.title = element_text(size = 16),
        legend.title.align = 0.05, 
        legend.title = element_text(size = 16), 
        legend.text = element_text(size = 14))
#p90

@@

```{r b95 }

@@

p95 <- ggplot()+
  geom_tile(data = sims, aes(x = d, y = d_h,
                        fill = as.factor(sig.n)),
            color="black")+
  facet_grid(h~a, labeller = label_both)+
  scale_x_continuous(breaks = c(-3:3), expand = c(0,0))+
  scale_y_continuous(breaks = c(-3:3), expand = c(0,0))+
  xlab("Current b")+
  ylab("Historical b")+
  scale_fill_manual(values = c("white", "blue"))+
  coord_equal()+
  theme_classic()+  
  labs(fill = "coverage") +
  theme(axis.text = element_text(size =10), 
        axis.title = element_text(size = 16), 
        legend.title.align = 0.05, 
        legend.title = element_text(size = 16), 
        legend.text = element_text(size = 14))

#p95

@@

For instance, a 90\% coverage rate means that in 900 of 1000 model fits, a given credible interval contained what we fixed as the value of current $b$. Put another way, if we set $b>0$, we calculate the proportion of the 90\% credible intervals that lie to the right of zero among the 1000 model fits.

The following figures depict the coverage rates of $b$ under an 80\%, 90\%, 95\% credible interval, respectively. In these figures, _historical b_ denotes the bias parameter for the generated historical strike data, and current $b$ means the bias parameter for the current data. The values on the right side represent the size of the historical data. The values on the top of the plots represent the power prior weight parameter $a$. For these figures, a darker color denotes a higher coverage rate, and vice versa.

```{r c80, fig.cap="80% credible interval"}

c80plot <- ggplot()+
  geom_tile(data = sims, aes(x = d, y = d_h,fill = C80),
            color="black")+
  facet_grid(h~a)+
   scale_fill_gradient2('C80', 
                        low = "darkred", mid = "white", 
                        high = "blue", 
                        breaks = c(0,0.5,0.7,0.9, 1), 
                        limits = c(0,1),
           labels = c("low", "", "moderate", "","high"))+
  scale_x_continuous(breaks = c(-3:3), expand = c(0,0))+
  scale_y_continuous(breaks = c(-3:3), expand = c(0,0))+
  xlab("Current b")+
  ylab("Historical b")+
  coord_equal()+
  theme_classic()+
  theme(legend.pos = "none",
        axis.text = element_text(size =10), 
        axis.title = element_text(size = 16), 
        #legend.title.align = 0.05, 
        #legend.title = element_text(size = 16), 
        #legend.text = element_text(size = 14)
        )

c80plot

@@

Figure 1 depicts the coverage rate with an 80\% credible interval. As the first column ($a=0.1$) depicts, when we place low weight on the historical strike data, we mostly rely on current information. When the bias parameter is 3, 2 or -2, -3, the coverage rate is high, very close to 1.

However, when the bias parameter ("current b") is 1, the coverage rate drops below 0.8. If the bias is not strong, our model can detect it with a lower accuracy. However, as we place more weight on the historical strike data (as $a$ increases), when the historical strike data is consistent with current information, the accuracy of detecting the bias is also close to 1 when the bias parameter is 1. When the bias parameter is 0 (no bias), however, the coverage rate is also around 80\%. Thus, if we use the 80\% credible interval, we have around 20\% chance of erroneously detecting non-zero bias.

```{r c90fig, fig.cap="90% credible interval"}

c90plot <- ggplot()+
  geom_tile(data = sims, aes(x = d, y = d_h,fill = C90),
            color="black")+
  facet_grid(h~a)+
   scale_fill_gradient2('C90', 
                        low = "darkred", mid = "white", 
                        high = "blue", 
                        breaks = c(0,0.5,0.7,0.9, 1), 
                        limits = c(0,1),
           labels = c("low", "", "moderate", "","high"))+
  scale_x_continuous(breaks = c(-3:3), expand = c(0,0))+
  scale_y_continuous(breaks = c(-3:3), expand = c(0,0))+
  xlab("Current b")+
  ylab("Historical b")+
  coord_equal()+
  theme_classic()+
  theme(legend.pos = "none",
        axis.text = element_text(size =10), 
        axis.title = element_text(size = 16), 
        #legend.title.align = 0.05, 
        #legend.title = element_text(size = 16), 
        #legend.text = element_text(size = 14)
        )

c90plot

Using a wider credible interval lowers this chance of error. For example, with a 90\% credible interval (Figure 2), when $b = 2$, the color is lighter (lower coverage rate) than with a 80\% credible interval; when $b=0$, the color is darker (higher coverage rate). The tradeoff is lower accuracy. With a 90\% credible interval, when the bias parameter is 2, the model is less accurate in detecting bias. When the bias parameter is 3, however, the coverage rate is still high.

```{r c95, fig.cap="95% credible interval"}
c95plot <- ggplot()+
  geom_tile(data = sims, aes(x = d, y = d_h,fill = C95),
            color="black")+
  facet_grid(h~a)+
   scale_fill_gradient2('C95', 
                        low = "darkred", mid = "white", 
                        high = "blue", 
                        breaks = c(0,0.5,0.7,0.9, 1), 
                        limits = c(0,1),
           labels = c("low", "", "moderate", "","high"))+
  scale_x_continuous(breaks = c(-3:3), expand = c(0,0))+
  scale_y_continuous(breaks = c(-3:3), expand = c(0,0))+
  xlab("Current b")+
  ylab("Historical b")+
  coord_equal()+
  theme_classic()+
  theme(legend.pos = "none",
        axis.text = element_text(size =10), 
        axis.title = element_text(size = 16), 
        #legend.title.align = 0.05, 
        #legend.title = element_text(size = 16), 
        #legend.text = element_text(size = 14)
        )

@@

Figure 3 depicts coverage rates when using a 95\% credible interval. Here, given $b = 0$, the coverage rate is very close to 1. This implies that we are very unlikely to conclude a bias when there is none. When the bias parameter is 3, the coverage rate is still high. This implies that we can still accurately detect the bias when it is strong. However, when the bias is moderate ($b = \{2, -2\}$, the coverage rate is only high when we have consistent historical information.

@@

```{r simltns, fig.align='center', fig.cap="Shuffled Sequence Simulation", out.width='50%'}

@@

b3.mean <-data.frame(b_s = mean(b3.t$b_s), 
                     `2.5%` =  mean(b3.t$`2.5%`),
                     `97.5%` = mean(b3.t$`97.5%`),
                     sim = "mean", 
                     b = "3",
                     sig = "Bias", # FIX: code as function
                     b.num = 3)

@@

The blue points represent the estimated posterior mean for the  bias parameter and the blue lines represent the 95\% credible interval. The red vertical line indicates the true value of the bias parameter.

@@

We describe here a prototype software application ("app") that implements the approach described above and that attorneys and others can use in real time to detect bias in the use of peremptory challenges. We built this app with `r R.version$version.string` and shiny [@R; @shiny].

@@

Statistical methods for _Batson_ and similar changes aim to help measure how much prospective jurors' race, gender, or other cognizable-class membership matters to how likely an attorney will use peremptory strikes against them. The Bayesian approach here accounts for the extent of the available historical strike data when estimating the posterior for the bias parameter. By incorporating the power prior, one can easily assess how sensitive that posterior estimate is to the available historical strike data. And with a tool like our prototype app, attorneys can use historical strike data to detect bias in real time during jury selection, thus making it easier to raise credible _Batson_ challenges.

@@

Finally, in practice, historical strike data tends to be incomplete at best, because jury-selection records in past cases are hard to get due to court practices and other barriers [@Grosso2017; @Wright2018]. For example, some jurisdictions do not require recording relevant or detailed enough information about prospective jurors who were or could have been struck (e.g., their race or gender). This forces researchers to turn to other data sources, if any, to find that information. And judges and court staff vary in how often they fully complete or maintain the information they are required to record.  This presents an issue for the approach presented here for detecting bias if the missing strike data is missing not at random.

>>>>

---
title: "Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data"
date: "`r Sys.Date()`"
author:
  - Sachin S. Pandya
#    orcid_id: 0000-0001-7387-1307
  - Xiaomeng Li
  - Timothy E. Moore
#    orcid_id: 0000-0002-9576-0517
#output: distill::distill_article
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: true
    fig_width: 5
    fig_height: 4
header-includes: 
  - \usepackage{amsmath}
fontsize: 12pt
bibliography: references.bib
abstract: |
  US law bars using peremptory strikes during jury selection because of prospective juror race, ethnicity, gender, or membership in certain other cognizable classes. Here, we propose a Bayesian approach for detecting such illegal bias. We show how to use historical data on an attorney's use of peremptory strikes in past cases to estimate, in a current trial, how likely an attorney has struck prospective jurors because of cognizable-class membership. We also use the _power prior_ to adjust the weight of such historical data in the analysis. We show how well our model performs on simulated data. Finally, we extend this approach with a prototype software application to detect bias in peremptory-challenge in real time during jury-selection. We illustrate this application's use with actual historical strike data from a convenience sample of cases from one federal district court.

@@

In any case, evidence of such illegal bias may include data on a party's use of peremptory challenges in past cases. (_Flowers v. Mississippi_ -@flowers2019, p. 2243; Wash. Rule 37(g)(v); Calif. Code of Civil Procedure $\S$ 231.7(d)(3)(G)). Prior studies have collected historical strike data in past cases and reported the observed difference in strike rates by race or gender. Typically, they test for the probability of observing a non-zero difference in strike rates by the race or gender of the struck prospective jurors, given repeated sampling from a hypothetical population of peremptory strikes with zero such difference [e.g., @Eisenberg2017; @Grosso2012; for discussion, see @Gastwirth2014; @Gastwirth2013]. Prior studies have also modeled how much a prospective juror's race affected the odds of being struck. For different modeling approaches using the same historical strike data from Mississippi, see @Craft2018a; @DeCamp2021; and @Dunn2021.

In this paper, we extend a Bayesian approach to _Batson_ and similar challenges [@Kadane2018] to incorporate historical strike data. In this approach, we specify a model of the peremptory-strike process in the court of interest that includes a bias parameter to which we assign an initial prior distribution. Then, we use an attorney's strike data from past trials, as generated by that same peremptory-strike process, to estimate a posterior distribution for that bias parameter. In so doing, we use the power prior [@Ibrahim2000; @Ibrahim2015] to control how much the strike history from past cases affects the posterior distribution of the bias parameter.  Finally, we use that posterior distribution as the prior for the bias parameter in the present case and update the posterior based on data on the target attorney's strikes in that present case.

@@

\begin{equation}
L(b|\delta) \propto L(b|\delta)(L(b|\delta_0))^{a_0}\exp(-\frac{b^2}{8})
\label{eq:model4}
\end{equation}

@@

To do this, we generated strike data for a current trial using seven different values of the bias parameter ($b_{curr} = \{-3,-2,-1,0,1,2,3\}$. As introduced above, $b>0$ ($b<0$) represents bias against (for) a prospective juror within (outside) the cognizable class, while $b=0$ denotes no bias.

We also generated historical strike data of three sizes (same, double, and triple the size of current data, i.e., data on one previous trial, two previous trials) using seven different values of the bias parameter for generating that historical data ($b_{hist} = \{-3,-2,-1,0,1,2,3\}$).

We let $b_{hist}$ differ from $b_{curr}$ (our parameter of interest) to simulate noise associated with historical strike data that is incomplete not at random. Alternatively, this difference simulates a real difference in the bias parameter that depends on a feature that the current trial has but that all the past trials in the historical data do not. To illustrate, suppose prosecutor with a bias value conditional on defendant race: $b > 0$ if the criminal defendant is Black, $b = 0$ if not. If our historical strike data contains only in cases with non-Black defendants, accounting for that historical data may lead us to underestimate prosecutor's bias in the current case ($b_{curr}$) with a Black defendant.

@@

We considered 735 scenarios in total, i.e., the combination of 7 different bias parameters of current data and of historical data, 3 different amounts of historical data and 5 different values of the power prior weight parameter $a$. For each scenario, we generated 1000 datasets and fit the model on those datasets. We recorded the posterior mean and credible intervals of one thousand model fits for each scenario.

In turn, for three credible interval levels (80\%, 90\% and 95\%), we recorded the coverage rates, i.e., the proportion of times we can reach the correct conclusion about the bias parameter based on those credible intervals. That is, when the true value of $b_{curr}$ is greater than zero, the credible interval lies to the right of zero. When $b_{curr}$ equals to zero, the credible interval includes zero, and when $b_{curr}$ is less than zero, the credible interval lies to the left of zero. Since we use a weakly-informative prior that pulls the estimate of $b_{current}$ towards zero, we focus not how accurately the model can recover the true value of $b_{curr}$, but on how accurately our model can detect that bias.

```{r b80, eval = FALSE}

@@

@@

```{r b90, eval = FALSE}

@@

@@

```{r b95, eval = FALSE}

@@

@@

For instance, a 90\% coverage rate means that in 900 of 1000 model fits, a given credible interval contained what we fixed as the value of $b_{curr}$. Put another way, if we set $b_{curr}>0$, we calculate the proportion of the 90\% credible intervals that lie to the right of zero among the 1000 model fits.

Figure 1 depicts the coverage rates of $b_{curr}$ with an 80\%, 90\%, and 95\% credible interval, respectively. Here, $b_{hist}$ denotes the bias parameter for the generated historical strike data and $b_{curr}$ denotes the bias parameter for the current data. The values on the right side represent the size of the historical data. The values on the top of the plots represent the power prior weight parameter $a$. For any given combination, a darker color shade denotes a higher coverage rate, and vice versa.

```{r c80}

library(latex2exp)
xlabel = TeX("$b_{curr}$")
ylabel = TeX("$b_{hist}$")

source(here::here("analysis","R","coverageplot_func.R"))

C80plot <- coverageplot_func(sims, sims$C80, xlabel, ylabel) +
           labs(title = "(a) Coverage Rates - 80% Credible Interval")

C90plot <- coverageplot_func(sims, sims$C90, xlabel, ylabel) +
           labs(title = "(b) Coverage Rates - 90% Credible Interval")

C95plot <- coverageplot_func(sims, sims$C95, xlabel, ylabel) +
           labs(title = "(c) Coverage Rates - 95% Credible Interval")

@@

```{r fig1a}
c80plot
```

```{r fig1b}
c90plot
```

```{r fig1c}
##| fig.cap=fig1caption

@@

Figure 1(a) depicts the coverage rate with an 80\% credible interval. As the first column ($a = 0.1$) depicts, when we place low weight on the historical strike data, we mostly rely on current information. When $b_{current}$ is 3, 2 or -2, -3, the coverage rate is high, very close to 1.

However, when $b_{curr}$ is 1, the coverage rate drops below 0.8. If the bias is not strong, our model can detect it with a lower accuracy. However, as we place more weight on the historical strike data (as $a$ increases), when the historical strike data is consistent with current information, the accuracy of detecting the bias is also close to 1 when the bias parameter is 1. When the bias parameter is 0 (no bias), however, the coverage rate is also around 80\%. Thus, if we use the 80\% credible interval, we have about a 20\% chance of erroneously detecting non-zero bias.

Using a wider credible interval lowers this chance of error. For example, with a 90\% credible interval (Figure 1(b)), when $b = 2$, the color is lighter (lower coverage rate) than with a 80\% credible interval; when $b=0$, the color is darker (higher coverage rate). The tradeoff is lower accuracy. With a 90\% credible interval, when the bias parameter is 2, the model is less accurate in detecting bias. When the bias parameter is 3, however, the coverage rate is still high.

Figure 1(c) depicts coverage rates when using a 95\% credible interval. Here, given $b_{curr} = 0$, the coverage rate is very close to 1. This implies that we are very unlikely to infer bias when there is none. When $b_{curr} = 3$, the coverage rate is still high. This implies that we can still accurately detect the bias when it is strong. However, when the bias is moderate ($b_{curr} = \{2, -2\}$), the coverage rate is only high when we have consistent historical information.

@@

```{r simltns}

#| fig.cap="Simulation results for simultaneous strikes example. Columns are values of bias used to simulate data. Blue points denote estimated posterior mean for the bias parameter. Blue lines denote the 95% credible interval. Red line denotes true value of bias parameters used in simulations."

@@

b3.mean <-data.frame(b_s = mean(b3.t$b_s), 
                     `2.5%` =  mean(b3.t$`2.5%`),
                     `97.5%` = mean(b3.t$`97.5%`),
                     sim = "mean", 
                     b = "3",
                     sig = "Bias",
                     b.num = 3)

@@

@@

We describe here a prototype software application ("app") that implements the approach described above and that attorneys and others can use in real time to detect bias in the use of peremptory challenges. We built this app with `r R.version$version.string` and Shiny [@R; @shiny].

@@

Statistical methods for _Batson_ and similar legal challenges have the purpose of helping to measure how much prospective jurors' race, gender, or other cognizable-class membership affected an attorney's use of peremptory strikes against them. The Bayesian approach here accounts for the extent of the available historical strike data when estimating the posterior for the bias parameter. By incorporating the power prior, one can easily assess how sensitive that posterior estimate is to the available historical strike data. And with a tool like our prototype app, attorneys can use historical strike data to detect bias in real time during jury selection, thus making it easier to raise credible _Batson_ challenges.

@@

Finally, in practice, historical strike data tends to be incomplete at best, because jury-selection records in past cases are hard to get due to court practices and other barriers [@Grosso2017; @Wright2018]. For example, some jurisdictions do not require recording relevant or detailed enough information about prospective jurors who were or could have been struck (e.g., their race or gender). This forces researchers to turn to other data sources, if any, to find that information. And judges and court staff vary in how often they fully complete or maintain the information they are required to record.  This presents an issue for the approach presented here if the missing strike data is missing not at random.
